---
title: "Cluster-Based Bouts"
author: "Paul R. Hibbing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cluster-Based Bouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PAutilities)
```

## Introduction and Installation

This vignette will show you how to run the cluster-based bout identifier. The
first step is making sure you have the `PAutilities` package (version 1.1.0 or
greater) installed on your computer. To get the most current code, you can
install from GitHub rather than CRAN. Here's how to do that:

```{r install, eval=FALSE}

## devtools is a package that makes it easy to install packages from GitHub, but
## I find that it sometimes struggles to install the related packages (i.e.,
## dependencies) correctly. So here is my manual workaround. All it's doing is
## looking through a list of required packages, and installing any of them that
## haven't already been installed (they'll be skipped if they have).
## Be aware: Some of these packages have long installation times.

invisible(lapply(
  c(
    "dplyr", "equivalence", "ggplot2", "graphics", "lazyeval",
    "lubridate", "magrittr", "methods", "reshape2", "rlang",
    "stats", "utils", "Rcpp"
  ),
  function(x) if (!x %in% installed.packages()) install.packages(x)
))

devtools::install_github("paulhibbing/PAutilities", dependencies = FALSE)

```

Copy and paste the above into your R console, then hit enter to run it.

## Preparation

Once you have the package installed, all you need is some activity data and the `get_bouts` function. For this demonstration, let's use some sample NHANES data.

```{r data, cache=TRUE}

## Find the file online
datafile <- file.path(
  "https://github.com/paulhibbing/PAutilities",
  "raw/master/data-raw/bouts_example.rds"
)

## Tell R where to save it
destination <- file.path(
  tempdir(),
  basename(datafile)
)

## Now save it
suppressMessages(
  utils::download.file(datafile, destination)
)

## Load the data
ex_data <- readRDS(destination)

## Discard the file now that we've loaded the data
invisible(file.remove(destination))

```

This dataset has activity counts that we can use to look at bouts of
moderate-to-vigorous physical activity (MVPA). For illustration, let's say we
initially coded our data as sedentary behavior (PAXINTEN &le; 100), light physical
activity (PAXINTEN 101-1951), or MVPA (PAXINTEN &ge; 1952).

```{r prep, cache=TRUE}

## Determine minute-by-minute intensity
intensity <- cut(
  ex_data$PAXINTEN,
  breaks = c(0, 101, 1952, Inf),
  labels = c("SB", "LPA", "MVPA"),
  right = FALSE
)

## The above command returns a factor variable -- the `get_bouts`
## function will complain about this and tell you it needs a
## character or numeric variable, so let's cast the intensity
## vector to character
intensity <- as.character(intensity)

```

## Running the Code

Once we have our data (`intensity` in this case), we can plug
it into `get_bouts`. Let's see the code first, then go over what it means.

```{r bouts, cache=TRUE}

mvpa_bouts <- PAutilities::get_bouts(
  x = intensity,
  method = "cluster-based",
  target = "MVPA",
  target_buffer = 30,
  longest_allowable_interruption = 2,
  required_percent = 80,
  max_n_interruptions = Inf,
  minimum_bout_length = 10
)

```

Here is what each piece means:

* **mvpa_bouts <-** Store the function results in an object called `mvpa_bouts`
* **PAutilities::get_bouts** This tells R to find the `get_bouts` function
  in the `PAutilities` package. In fact, if you run `PAutilities::get_bouts`
  in your console, R will print the source code.
* **x = intensity** Here we specify that our input datastream is `intensity`,
  as defined in the earlier code.
* **method = "cluster-based"** Here we specify that R should run the
  cluster-based bout identification method. Right now, this is the only
  option, but in the future it will be possible to add others.
* **target = "MVPA"** Here we specify which behavior we are interested in.
  The input data (`intensity`) has values in the set `{SB, LPA, MVPA}`, and
  we would like to look specifically at bouts of `MVPA`, with the other
  behaviors being lumped together in a single group called `other`.
* **target_buffer = 30** Here we specify how our data should be
  stratified/partitioned. In this case, `intensity` is a minute-by-minute
  variable, so our setting of `30` means the data will be
  stratified/partitioned anytime we see &ge;30 consecutive minutes of `other`
  behavior. *It is crucial to understand that the meaning of this setting
  depends on the epoch length of the input data. If `intensity` was a
  second-by-second variable, we would need to set `target_buffer = 1800` to
  achieve the same 30-min threshold we are using for this example.*
* **longest_allowable_interruption = 2** Here we specify that a valid bout
  should not include any single interruption lasting longer than 2 minutes.
  *(Again, this is dependent on epoch length; a setting of 120 would be
  needed to achieve the same threshold for a second-by-second input variable)*
* **required_percent = 80** Here we specify that a valid bout should be
  interrupted for no more than 20% of its full duration.
* **max_n_interruptions = Inf** Here we specify that a valid bout can have
  unlimited interruptions as long as the criteria for
  `longest_allowable_interruption` and `required_percent` are met.
* **minimum_bout_length = 10** Here we specify that only bouts lasting &ge; 10
  min should be included in the output.
  
The above elements are set up to allow flexible bout criteria depending on the
research question and the variable of interest. In our example, we set
`max_n_interruptions = Inf` to avoid a restriction in that area -- similar
approaches can be taken for other settings as well, by setting them to `0` or
`Inf` as appropriate. (For `required_percent`, 100 is the upper limit rather
than `Inf`.) Notably, `minimum_bout_length` is a filtering criterion. It has no
direct affect on how the bouts are defined; it simply affects which ones are
retained *after* they have been defined.
  
## Interpreting the Output

Now let's take a look at the output and go over what it means:

```{r output, cache=TRUE}
mvpa_bouts
```

This is a data frame with one row per bout. The variables are:

* **start_index** The starting point of the bout (e.g., `intensity[3809]` for
  the first bout in this example)
* **end_index** The ending point of the bout (e.g., `intensity[3828]` for
  the first bout in this example)
* **values** A meaningless constant (equal to the setting of `target`),
  left over from run length encoding
* **n_total_events** The number of distinct behavior events occurring between
  `start_index` and `end_index`
* **n_value_events** The number of distinct target behavior events occurring
  between `start_index` and `end_index` (referred to as `value` events in
  reference to the `values` column)
* **n_interruption_events** The number of distinct interruption events
  occurring between `start_index` and `end_index`
* **length_total** The combined duration of all `value`
  and `interruption` events
* **length_value** The combined duration of all `value` events
* **length_interruption** The combined duration of all `interruption` events
* **longest_interruption_event** The duration of the single longest
  interruption event
* **percent_time_engaged** Percentage of `length_total` comprised
  by `length_value`
  
As before, epoch length influences the interpretation of length variables. Keep that in mind.

## Expanding the Output

Lastly, in some cases we may want to convert our bout information back to the original length of the input (i.e., `intensity`). We can use the `bout_expand` function to accomplish that.

```{r expansion, cache=TRUE}
expanded <- PAutilities::bout_expand(mvpa_bouts)
str(expanded)
```

And we can also append that new variable into our original dataset as well.

```{r expansion2, cache=TRUE}
ex_data$intensity <- expanded
head(ex_data)
```

## Conclusion

This should give you a broad sense of how to use the cluster-based bout identification method and what else you can do with it. Feel free to [post an issue on the GitHub page](https://github.com/paulhibbing/PAutilities/issues/new/choose) if any of the above gives you trouble. Good luck!
